A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-23 tex/body.tex
more
minor

{\looseness=-1
Datatypes and codatatypes are an integral part of many proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, datatypes
have emerged in a few automatic theorem provers as well. The SMT-LIB
%\cite{barrett-et-al-2010}
format, implemented by most SMT
%(satisfiability modulo theories)
solvers, has been extended with a syntax for datatypes.
In this \thewordpaper, we introduce a
unified \hfill decision \hfill procedure \hfill for \hfill universal \hfill problems \hfill involving \hfill datatypes \hfill and \hfill codatatypes%
\begin{conf}\goodbreak\noindent\end{conf}%  %%% TYPESETTING: terrible HACK
in combination (Section~\ref{sec:the-decision-procedure}).
The procedure is described abstractly as a %tableau-like
calculus and can be
composed via the Nelson--Oppen method \cite{nelson-oppen-1979}.
It generalizes the procedure by Barrett et al.\ \cite{barrett-et-al-2007}, which covers
only datatypes.
%To our knowledge, our procedure is the first of its kind for the theory of
%codatatypes.
\begin{conf}
Detailed proofs %of completeness
are included in a %technical
report
\cite{our-report}.
\end{conf}%

@@

Before constructing a model for $F\!$, the theory solver constructs neither $\mu$-terms nor the mapping $\Val$.
Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
To detect whether \rn{Acyclic} applies,
the procedure considers each equivalence class $\ec{t}$ containing a datatype constructor $\const{C}( t_1, \ldots, t_n )$.
It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all constructor arguments in these equivalence classes recursively.
If while doing so it returns to $\ec{t}$, it deduces that \rn{Acyclic} applies. % to $t$.
To recognize when the precondition of \rn{Unique} holds,
the procedure considers the set $\SSS$ of all codatatype equivalence classes. It
\begin{conf}\goodbreak\noindent\end{conf}%
simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $\SSS$,
while partitioning $\SSS$ into $\SSS_1, \ldots, \SSS_n$ based on the top-most symbol of constructor terms in these equivalence classes
and the equivalence of their arguments of ordinary types.
It then partitions each set %in this partition %in the same manner
recursively.
If %after doing so,
the resulting partition contains a set $\SSS_i$ containing distinct terms $u$
and $v$, it deduces that \rn{Unique} applies to $u$ and~$v$.

@@

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the theory files from the Isabelle distribution (Distro, 879~goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. \hfill We~also%
\begin{conf}%
\par
%%% TYPESETTING: Terrible hack
\pagebreak
\noindent
\end{conf}%
added two unpublished theories about Bird and Stern--Brocot trees
by Peter Gammie and Andreas Lochbihler (\gandl,
317~goals). To exercise the datatype
support, the benchmarks were complemented by theories
about lists and trees. The theories were selected before conducting the
experiments. The experimental data are publicly available.%
\footnote{\url{http://lara.epfl.ch/~reynolds/CADE2015-cdt/}}
%http://www21.in.tum.de/~blanchet/codp_data.tgz}}

>>>>

{\looseness=-1
Datatypes and codatatypes are an integral part of many proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, datatypes
have emerged in a few automatic theorem provers as well. The SMT-LIB
%\cite{barrett-et-al-2010}
format, implemented by most SMT
%(satisfiability modulo theories)
solvers, has been extended with a syntax for datatypes.
In this \thewordpaper, we introduce a
unified \hfill decision \hfill procedure \hfill for \hfill universal \hfill problems \hfill involving \hfill datatypes \hfill and \hfill codatatypes%
\begin{conf}\goodbreak\noindent\end{conf}%  %%% TYPESETTING BREAK
\begin{rep} \end{rep}%
in combination (Section~\ref{sec:the-decision-procedure}).
The procedure is described abstractly as a %tableau-like
calculus and can be
composed via the Nelson--Oppen method \cite{nelson-oppen-1979}.
It generalizes the procedure by Barrett et al.\ \cite{barrett-et-al-2007}, which covers
only datatypes.
%To our knowledge, our procedure is the first of its kind for the theory of
%codatatypes.
\begin{conf}
Detailed proofs %of completeness
are included in a %technical
report
\cite{our-report}.
\end{conf}%

@@

Before constructing a model for $F\!$, the theory solver constructs neither $\mu$-terms nor the mapping $\Val$.
Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
To detect whether \rn{Acyclic} applies,
the procedure considers each equivalence class $\ec{t}$ containing a datatype constructor $\const{C}( t_1, \ldots, t_n )$.
It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all constructor arguments in these equivalence classes recursively.
If while doing so it returns to $\ec{t}$, it deduces that \rn{Acyclic} applies. % to $t$.
To recognize when the precondition of \rn{Unique} holds,
the procedure considers the set $\SSS$ of all codatatype equivalence classes. It
\begin{conf}\goodbreak\noindent\end{conf}%  %%% TYPESETTING BREAK
simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $\SSS$,
while partitioning $\SSS$ into $\SSS_1, \ldots, \SSS_n$ based on the top-most symbol of constructor terms in these equivalence classes
and the equivalence of their arguments of ordinary types.
It then partitions each set %in this partition %in the same manner
recursively.
If %after doing so,
the resulting partition contains a set $\SSS_i$ containing distinct terms $u$
and $v$, it deduces that \rn{Unique} applies to $u$ and~$v$.

@@

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the theory files from the Isabelle distribution (Distro, 879~goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We
added two unpublished theories about Bird and Stern--Brocot trees
by Peter Gammie and Andreas Lochbihler (\gandl,
317~goals). To exercise the datatype
support, the benchmarks were complemented by theories
about lists and trees. The theories were selected before conducting the
experiments. The experimental data are publicly available.%
\footnote{\url{http://lara.epfl.ch/~reynolds/CADE2015-cdt/}}
%http://www21.in.tum.de/~blanchet/codp_data.tgz}}
